Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-14 paper.tex
update discussion

For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
structures.
Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
under Hudson. \citep{Wong_a-general_2023}

@@

Because the likelihood is computed based on edges,
and edges are no longer recorded once the local most recent common ancestor 
has been reached, we will underestimate the number of positions where 
a recombination could have occurred in case a recombination hits a 
lineage with trapped non-ancestral material (see end \ref{par:description}). 
If we assume that only a single recombination 
occurred, then it suffices however to provide a correction factor $g$ in equation
\ref{eq:depth}. Here $g$ represents the 
length of non-ancestral material along which a recombination would have resulted  
in the same observed local sequence of tree changes.

\begin{figure}[!ht] \label{fig:algo}
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Description of algorithm: Moving along the genome we transition from the first 
local tree (1) to the next (4). For each edge that is removed from the first tree we 
decrement the count in the corresponding internode intervals and update the identity of 
the last parent. Subsequently, for each edge that will be inserted we compute the likelihood 
associated with that edge prior to incrementing the counts.}
\end{figure}

@@

% ADD IN RESULTS
Figure \ref{sup:fig:vs-hudson} shows the correlation between the likelihood under the hudson 
for an ARG with detailed recombination information simulated and computed using 
\texttt{msprime} \citep{baumdicker_efficient_2021}. Show impact of unary nodes on $r^2$.
Simulated mutation data under the SMC and passed this to ARGweaver.
Scatterplot of coalescent prior as returned by ARGweaver \citep{rasmussen_genome-wide_2014} 
for a 100 samples from the posterior against the likelihood as described here fore these 
sampled ARGs \ref{sup:fig:vs-argweaver}

Implementation of the algorithm is available on https://github.com/GertjanBisschop/runsmc.

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \texttt{tskit} library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with.

Our approach therefore strenghtens the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome.

% Algorithm is suitable for a general class of ARGs each with their own specific features.
% What happens in the case of ARGs that do not meet our assumptions: absent unary nodes, polytomies, ...
The above algorithm treats each edge independently which implies we can deal with polytomies.
Furthermore, although we rely heavily on unary nodes to provide us with the necessary 
information on recombination events, our algorithm can also accomodate for 
those inferred ARGs that lack unary nodes (e.g.\ \relate, and see figure in supplements).
However, in case of (sub)tree height changes along the ARG, the algorithm would infer 
two recombination events along the lineages associated 
with both edges switching parent nodes. This can be mitigated by limiting the number of 
'inferrable' recombination events per coalescence event to 1 (see supplementary figure).

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and a set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods combining the best of both 
heuristic and model based inference approaches. Thirdly, quantify uncertainty.

1/ statistical inference
This is an important step towards bridging the gap between the tremendous 
recent advances in ARG inference and the current state of the art statistical inference of 
past demography (but see ): currently most/all methods perform ARG-based analysis 
by extracting mariginal trees from the ARG \citep{hejase_2022} % likelihood-based demographic inference on local trees (ref?) 
The provided theory and associated algorithm will function as a building block of a 
demography-aware genealogy sampler. Current approach couple of major advantages:
* any type of inferred ARG as a starting point. Potentially implying a massive reduction in 
MCMC runtime by reducing the necessary burnin period.
* by integrating out the exact time to the recombination event we further restrict the size 
of ARG space that needs to be explored.
* exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
* we can readily incorporate $N_e$-changes as well as multiple populations and migration.\\

Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
method. Once number of lineages has reduced sufficiently, model-based approach can be used. One 
of the advantages of formulating this backwards in time. 
Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}\\

Both of these future directions tie in with the ability to quantify uncertainty by sampling 
from an underlying distribution. This will in particular enable us to deal with those 
parts in both time and along the genome for which we have insufficient information to confidently 
reconstruct the ARG. For example the presence of polytomies in \tsinfer represents 
uncertainty. Systematicallty break them by sampling from the coalescent rather then randomly.
Quantifying uncertainty would not necessarily always have to be associated with dealing with 
the entire subset of all compatible genealogies. Uncertainty could also be quantified and passed 
on as metadata to a parent node representing the uncertainty in the interval spanned by the 
polytomy.

@@

% final conclusion

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.5\textwidth]{example-image-a}
\caption{Compare likelihoods inferred ARGs ARGweaver}
 \label{sup:fig:vs-argweaver}
\end{figure}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.5\textwidth]{example-image-a}
\caption{Compare likelihoods Hudson and SMC}
\label{sup:fig:vs-hudson}
\end{figure}

>>>>

For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
structures.
Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
under Hudson. \citep{wong_a-general_2023}

@@

Because the likelihood is computed based on edges,
and edges are no longer recorded once the local most recent common ancestor 
has been reached, we will underestimate the number of positions where 
a recombination could have occurred in case a recombination hits a 
lineage with trapped non-ancestral material (see end \ref{par:description}). 
If we assume that only a single recombination could have  
occurred, then it suffices however to provide a correction factor $g$ in equation
\ref{eq:depth}. Here $g$ represents the 
length of non-ancestral material along which a recombination would have resulted  
in the same observed local sequence of tree changes.

\begin{figure}[!ht] \label{fig:algo}
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Description of algorithm: Moving along the genome we transition from  
local tree $T_1$ (1) to the next, $T_2$ (4). Edges that do not persist in $T_2$ are 
removed (2). Edges starting at the left coordinate of $T_2$ are inserted (3). 
After deletion or insertion of an edge the count vector $I$ is updated along the 
corresponding internode intervals. Prior to updating $I$ following an insertion, 
the likelihood for that edge is computed as $I$ then represents the total number of 
lineages the lineage associated with this new edge can 
coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
\end{figure}

@@

% ADD IN RESULTS
Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation 
between the coalescent prior as returned by \argweaver and the likelihood as computed here. 
We simulated 1Mb of data for 10 diploid individuals 
under the SMC using human-like parameters. 100 samples were taking from the posterior every for 
a total of 1000 MCMC iterations.

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \texttt{tskit} library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with. Our approach therefore strenghtens 
the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome by respectively adding intervals to $I$ 
or by splitting edges when/wherever the parameter set changes.
[NOTE: these options have already been implemented in the current implementation]

Although we rely heavily on unary nodes to provide us with the necessary 
information on recombination events, our algorithm only implicitly assumes there presence.
The algorithm can therefore also accomodate for 
those inferred ARGs that lack unary nodes (e.g.\ \relate). 
Note that without unary nodes, our algorithm will detect a recombination event along 
both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
edges would switch parent nodes. This can be mitigated by limiting the number of 
'inferrable' recombination events per coalescence event to 1.
% simulated results
We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
After removing the recombination nodes the likelihoods as defined here were computed both with and 
without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
explained is higher when the unary nodes are retained.

% polytomies
[TO DO: polytomies] The above algorithm treats each edge independently which implies 
we can deal with polytomies.

\section{Discussion}

Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and an associated set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods that combine the best of both 
heuristic and model-based inference approaches are now possible.
Thirdly, using a model we can quantify the uncertainty inherently associated 
with ARG-reconstruction.

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of a 
demography-aware genealogy sampler that would allow us to incorporate the information 
provided by recombination events.
More concretely, given our algorithm, any inferred ARG can be used a good starting point 
for a future MCMC-sampler. This potentially implies a massive runtime reduction by 
reducing the necessary burn-in period. Evenmore, by integrating out the exact time 
to the recombination event we further restrict the size of the ARG space that needs to 
be explored. 
This algorithm will also allow us to exploit the efficiencies of the tree sequence 
data structure (see \citep{mahmoudi_bayesian_2022}) to evaluate any new proposal.
We can readily incorporate $N_e$-changes as well as multiple populations and migration.\\

% 2/ hybrid ARG-inference
Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
method. Once number of lineages has reduced sufficiently, model-based approach can be used. One 
of the advantages of formulating this backwards in time. 
Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}\\

% 3/ quantifying uncertainty
Both of these future directions tie in with the ability to quantify uncertainty by sampling 
from an underlying distribution. This will in particular enable us to deal with those 
parts in both time and along the genome for which we have insufficient information to confidently 
reconstruct the ARG. For example the presence of polytomies in \tsinfer represents 
uncertainty. Systematicallty break them by sampling from the coalescent rather then randomly.
Quantifying uncertainty would not necessarily always have to be associated with dealing with 
the entire subset of all compatible genealogies. Uncertainty could also be quantified and passed 
on as metadata to a parent node representing the uncertainty in the interval spanned by the 
polytomy.

@@

% final conclusion
\section{Data availability}

All scripts and data used for this manuscript are available on https://github.com/gertjanbisschop/smc-bit-paper
An implementation of the algorithm is available on https://github.com/gertjanbisschop/runsmc.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/argweaver_vs_runsmc.png}
\caption{Compare likelihoods inferred ARGs ARGweaver: placeholder for final figure with similar content}
 \label{sup:fig:vs-argweaver}
\end{figure}

\begin{figure}[!ht]
\centering
\includegraphics[width=0.75\textwidth]{figures/supplementary-figs/v_hudson_unary_simpl.png}
\caption{Compare likelihoods Hudson and SMC with and without unary nodes: placeholder for final figure with similar content}
\label{sup:fig:vs-hudson}
\end{figure}
